Examining background gene lists using simulation - comparison of different enrichment techniques

### 2024-09-10 analysis/sim/run_simpw.Rmd
sim update for granular data viz with vioplots boxplots etc

SIMS=1000
#run 500 sims
FRAC_DE=0.05
FC=0.5
N_REPS=3
DGE_FUNC="deseq2"
SUM_COUNT=2e7
VARIANCE=c(0,0.1,0.2,0.3,0.4,0.5,0.6)

@@

res <- lapply(1:nrow(mygrid), function(i) {
  FRAC_DE=mygrid[i,"FRAC_DE"]
  FC=mygrid[i,"FC"]
  N_REPS=mygrid[i,"N_REPS"]
  DGE_FUNC=as.character(mygrid[i,"DGE_FUNC"])
  SUM_COUNT=mygrid[i,"SUM_COUNT"]
  VARIANCE=mygrid[i,"VARIANCE"]
  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
  as.data.frame(do.call(rbind, x))
})

res <- do.call(rbind,res)

@@

@@

@@

>>>>

SIMS=50
#run 1000 sims
FRAC_DE=0.05
FC=0.5
N_REPS=3
DGE_FUNC="deseq2"
SUM_COUNT=2e7
VARIANCE=c(0,0.1,0.2,0.3,0.4,0.5,0.6)

@@

res <- lapply(1:nrow(mygrid), function(i) {
  FRAC_DE=mygrid[i,"FRAC_DE"]
  FC=mygrid[i,"FC"]
  N_REPS=mygrid[i,"N_REPS"]
  DGE_FUNC=as.character(mygrid[i,"DGE_FUNC"])
  SUM_COUNT=mygrid[i,"SUM_COUNT"]
  VARIANCE=mygrid[i,"VARIANCE"]
  agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=16)
})

@@

## Aggregate data for charting as vioplot

```{r,vioagg}

library(vioplot)
par(mfrow=c(3,3))

## clusterprofiler default
cp <- lapply(res,"[[",1)
cp2 <- lapply(cp,function(df) {
  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
  df$f <- 2*df$p*df$r/(df$p+df$r)
  df
} )

cp2p <- lapply(cp2,"[[","p")
cp2p <- lapply(cp2p,function(x) { x[is.na(x) ]<- 0 ; x } )

cp2r <- lapply(cp2,"[[","r")
cp2r <- lapply(cp2r,function(x) { x[is.na(x) ]<- 0 ; x } )

cp2f <- lapply(cp2,"[[","f")
cp2f <- lapply(cp2f,function(x) { x[is.na(x) ]<- 0 ; x } )

names(cp2p) <- names(cp2r) <- names(cp2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")

vioplot(cp2p,xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
vioplot(cp2r,xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
vioplot(cp2f,xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()

## clusterprofiler bgfix
cpbg <- lapply(res,"[[",2)
cpbg2 <- lapply(cpbg,function(df) {
  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
  df$f <- 2*df$p*df$r/(df$p+df$r)
  df
} )

cpbg2p <- lapply(cpbg2,"[[","p")
cpbg2p <- lapply(cpbg2p,function(x) { x[is.na(x) ]<- 0 ; x } )

cpbg2r <- lapply(cpbg2,"[[","r")
cpbg2r <- lapply(cpbg2r,function(x) { x[is.na(x) ]<- 0 ; x } )

cpbg2f <- lapply(cpbg2,"[[","f")
cpbg2f <- lapply(cpbg2f,function(x) { x[is.na(x) ]<- 0 ; x } )

names(cpbg2p) <- names(cpbg2r) <- names(cpbg2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")

vioplot(cpbg2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
vioplot(cpbg2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
vioplot(cpbg2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()

## clusterprofiler FDR fix
cpfdr <- lapply(res,"[[",3)
cpfdr2 <- lapply(cpfdr,function(df) {
  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
  df$f <- 2*df$p*df$r/(df$p+df$r)
  df
} )

cpfdr2p <- lapply(cpfdr2,"[[","p")
cpfdr2p <- lapply(cpfdr2p,function(x) { x[is.na(x) ]<- 0 ; x } )

cpfdr2r <- lapply(cpfdr2,"[[","r")
cpfdr2r <- lapply(cpfdr2r,function(x) { x[is.na(x) ]<- 0 ; x } )

cpfdr2f <- lapply(cpfdr2,"[[","f")
cpfdr2f <- lapply(cpfdr2f,function(x) { x[is.na(x) ]<- 0 ; x } )

names(cpfdr2p) <- names(cpfdr2r) <- names(cpfdr2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")

vioplot(cpfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
vioplot(cpfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
vioplot(cpfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()

## clusterprofiler BG and FDR fix
cpbgfdr <- lapply(res,"[[",4)
cpbgfdr2 <- lapply(cpbgfdr,function(df) {
  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
  df$f <- 2*df$p*df$r/(df$p+df$r)
  df
} )

cpbgfdr2p <- lapply(cpbgfdr2,"[[","p")
cpbgfdr2p <- lapply(cpbgfdr2p,function(x) { x[is.na(x) ]<- 0 ; x } )

cpbgfdr2r <- lapply(cpbgfdr2,"[[","r")
cpbgfdr2r <- lapply(cpbgfdr2r,function(x) { x[is.na(x) ]<- 0 ; x } )

cpbgfdr2f <- lapply(cpbgfdr2,"[[","f")
cpbgfdr2f <- lapply(cpbgfdr2f,function(x) { x[is.na(x) ]<- 0 ; x } )

names(cpbgfdr2p) <- names(cpbgfdr2r) <- names(cpbgfdr2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")

vioplot(cpbgfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(cpbgfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(cpbgfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()

## FORA
fo <- lapply(res,"[[",5)
fo2 <- lapply(fo,function(df) {
  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
  df$f <- 2*df$p*df$r/(df$p+df$r)
  df
} )

fo2p <- lapply(fo2,"[[","p")
fo2p <- lapply(fo2p,function(x) { x[is.na(x) ]<- 0 ; x } )

fo2r <- lapply(fo2,"[[","r")
fo2r <- lapply(fo2r,function(x) { x[is.na(x) ]<- 0 ; x } )

fo2f <- lapply(fo2,"[[","f")
fo2f <- lapply(fo2f,function(x) { x[is.na(x) ]<- 0 ; x } )

names(fo2p) <- names(fo2r) <- names(fo2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")

vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()

## FGSEA
fg <- lapply(res,"[[",6)
fg2 <- lapply(fg,function(df) {
  df$p <- df$true_pos/(df$true_pos+ df$false_pos)
  df$r <- df$true_pos/(df$true_pos+ df$false_neg)
  df$f <- 2*df$p*df$r/(df$p+df$r)
  df
} )

fg2p <- lapply(fg2,"[[","p")
fg2p <- lapply(fg2p,function(x) { x[is.na(x) ]<- 0 ; x } )

fg2r <- lapply(fg2,"[[","r")
fg2r <- lapply(fg2r,function(x) { x[is.na(x) ]<- 0 ; x } )

fg2f <- lapply(fg2,"[[","f")
fg2f <- lapply(fg2f,function(x) { x[is.na(x) ]<- 0 ; x } )

names(fg2p) <- names(fg2r) <- names(fg2f) <- c("0","0.1","0.2","0.3","0.4","0.5","0.6")

vioplot(fg2p,xlab="introduced noise",ylab="precision",main="fgsea") ; grid()
vioplot(fg2r,xlab="introduced noise",ylab="recall",main="fgsea") ; grid()
vioplot(fg2f,xlab="introduced noise",ylab="F1",main="fgsea") ; grid()

png("fig3_vio.png", width=7,height=7,units="in",res=150,pointsize=12)
par(mfrow=c(3,3))
vioplot(cp2p,xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
vioplot(cp2r,xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
vioplot(cp2f,xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
vioplot(cpbg2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
vioplot(cpbg2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
vioplot(cpbg2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
vioplot(cpfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
vioplot(cpfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
vioplot(cpfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
dev.off()
png("fig3_vio2.png", width=7,height=7,units="in",res=150,pointsize=12)
par(mfrow=c(3,3))
vioplot(cpbgfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(cpbgfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(cpbgfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
dev.off()

pdf("fig3_vio.pdf", width=7,height=7,pointsize=12)
par(mfrow=c(3,3))
vioplot(cp2p,xlab="introduced noise",ylab="precision",main="clusterProfiler") ; grid()
vioplot(cp2r,xlab="introduced noise",ylab="recall",main="clusterProfiler") ; grid()
vioplot(cp2f,xlab="introduced noise",ylab="F1",main="clusterProfiler") ; grid()
vioplot(cpbg2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG fix") ; grid()
vioplot(cpbg2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG fix") ; grid()
vioplot(cpbg2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG fix") ; grid()
vioplot(cpfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler FDR fix") ; grid()
vioplot(cpfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler FDR fix") ; grid()
vioplot(cpfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler FDR fix") ; grid()
vioplot(cpbgfdr2p,xlab="introduced noise",ylab="precision",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(cpbgfdr2r,xlab="introduced noise",ylab="recall",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(cpbgfdr2f,xlab="introduced noise",ylab="F1",main="clusterProfiler BG and FDR fix") ; grid()
vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
vioplot(fo2p,xlab="introduced noise",ylab="precision",main="fora") ; grid()
vioplot(fo2r,xlab="introduced noise",ylab="recall",main="fora") ; grid()
vioplot(fo2f,xlab="introduced noise",ylab="F1",main="fora") ; grid()
dev.off()

@@

res2 <- lapply(res,function(x) { lapply(x,function(y) { colMeans(y[,1:3]) } ) } )
res2 <- lapply(res2,function(x) {
  out <- as.data.frame(do.call(rbind,x))
  pw <- c("cp","cpbg","cpfdr","cpbgfdr","fora","fgsea")
  out$pw <- pw
  out
} )
res3 <- do.call(rbind,res2)
res3$var <- c(rep(0,6),rep(0.1,6),rep(0.2,6),rep(0.3,6),rep(0.4,6),rep(0.5,6),rep(0.6,6))

@@

######################################
# Now as boxplot
library(beeswarm)

prec <- list("CP default"=cp$p,"CP BG fix"=cpbg$p,"CP FDR fix"=cpfdr$p,"CP BG and FDR fix"=cpbgfdr$p,"fora"=fo$p,"fgsea"=fg$p)
rec <- list("CP default"=cp$r,"CP BG fix"=cpbg$r,"CP FDR fix"=cpfdr$r,"CP BG and FDR fix"=cpbgfdr$r,"fora"=fo$r,"fgsea"=fg$r)
f1 <- lapply(1:length(prec), function(i) { 2/(1/rec[[i]]+1/prec[[i]]) })
names(f1) <- names(prec)

par(mfrow=c(1,3))
par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
beeswarm(prec,add=TRUE,col="black",pch=19)
boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
beeswarm(rec,add=TRUE,col="black",pch=19)
boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
beeswarm(f1,add=TRUE,col="black",pch=19)

png("fig3_box.png", width=7,height=5,units="in",res=150,pointsize=12)
par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
par(mfrow=c(1,3))
boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
beeswarm(prec,add=TRUE,col="black",pch=19)
boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
beeswarm(rec,add=TRUE,col="black",pch=19)
boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
beeswarm(f1,add=TRUE,col="black",pch=19)
dev.off()

pdf("fig3_bars.pdf", width=7,height=5,pointsize=12)
par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
par(mfrow=c(1,3))
boxplot(prec,cex=0,main="precision",ylab="",col="white",las=2)
beeswarm(prec,add=TRUE,col="black",pch=19)
boxplot(rec,cex=0,main="recall",ylab="",col="white",las=2)
beeswarm(rec,add=TRUE,col="black",pch=19)
boxplot(f1,cex=0,main="F1 index",ylab="",col="white",las=2)
beeswarm(f1,add=TRUE,col="black",pch=19)
dev.off()
